### 2016-05-04 papers/thesis-vischer/Thesis.tex
more cleanup, revision, content to methods Tests skipped!

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\setlength{\parindent}{0cm} % don't indent new paragraphs...
%\parskip \psl % ... place a space between paragraphs instead
\usepackage{graphicx, float}
\usepackage[font=small, labelfont=bf]{caption}
%\usepackage[english]{babel}
\usepackage{tikz}
%\usepackage{biblatex}
\usepackage{amsmath}
\usepackage[inline]{enumitem}
%\usepackage{float}
\usetikzlibrary{automata, positioning}
\newcommand{\ignore}[1]{}
\newcommand*{\diff}{\mathrm{d}}
\newcommand{\some}{{\color{red} something}}
\renewcommand{\vec}[1]{\mathbf{#1}}

@@

\renewcommand{\title}{Application of Generalized Renormalization Group Methods to the Square-Well Liquid Near Criticality}

@@

\section{Introduction}
\subsection{Motivation}
Fluids have long garnered interest\cite{theorysimpleliquids}\cite{theoryofcrits}, particularly among computationally-minded researchers, as an area in which current models are unequipped to robustly predict behavior across multiple regimes. A largely troublesome \cite{white2001global} regime is near the so-called ``critical point'' where liquid and vapor are indistinguishable and, more importantly, large fluctuations of fluid density occur. Most traditional fluid models assume the density across the cell to be nearly constant. As a result, these methods tend to poorly handle these fluctuations near criticality. 
Using the well-established model of the square-well fluid, we seek to determine the applicability of a particular generalized renormalization-group formalism (``GRG'') to find the absolute Helmholtz free energy of systems near criticality.\\

Past efforts to incorporate renormalization group methods into standard Statistical Associating Fluid Theory (``SAFT'') methods have seen success, particularly the treatment offered by Forte et al.~\cite{forte2011application} with the inclusion of a fitted parameter related to the correlation length. GRG's history in fluid simulation has largely been a fitted one - that is, determining a free parameter and then simulating a vast number of cells, varying the parameter each time, to find the most well-suited value for a particular fluid. We present a non-fitted approach that is GRG-minded from the ground up. We exploit the periodicity of Monte-Carlo methods to decompose large cells in the thermodynamic limit into small cells that only allow certain fluctuation wavelengths. We then carefully couple simulations of different ``doubling regimes'', each of which considers a different maximum wavelength of fluctuations, where the number density remains fixed but the volume and number have increased. Summing the difference in absolute Helmholtz free energy $\Delta F^{\text{abs}}$ between doubling regimes will then approximate the absolute free energy of the full cell, including near criticality. Finding absolute free energies is a somewhat challenging problem, but is certainly surmountable with more Monte Carlo. \\
We will begin by outlining the necessary models and assumptions/approximations for this work. We will then detail the algorithms used to compute the quantities with the the discussion on results following. 
\subsection{Hard Sphere}
The hard sphere model approximates monoatomic fluids as having no interactions other than collisions. These spheres exhibit no attraction or repulsion, and consequently each carries kinetic energy $T = 3/2 k_b T$. There is also a harder-to-calculate ``excess'' energy to be considered (which, in fact, contains most of the interesting behavior) which will be discussed further on. The system is defined to have zero total linear momentum, and any collision between two or more spheres is perfectly elastic. For simulation purposes, the system is required to have periodic boundary conditions - that is, when a sphere would classically collide with the wall of the cell it instead breaches the wall and then reappears on the other side of the cell with the same velocity. \\
\ignore{The hard sphere model is incredibly useful; as high temperatures are approached, energy associated with sphere location tends to asymptote for most models [are there any that this isn't true for?] and the kinetic term consequently dominates.}

\subsection{Square-Well Liquid}
The square-well liquid is a ubiquitous approximation that sees extensive use throughout computational fluid research. This simple fluid model consists of hard spheres that are subject to an intermolecular pair potential - that is, an attractive force between spheres. This model consists of several key parameters: sphere size $\sigma$, well width $\lambda$, and well depth $\epsilon$. \\
 As a hard-sphere model, the spheres are not allowed to overlap. Any arrangement of spheres - hereafter referred to as the ``state'' of the system - that involves either two or more spheres overlapping or a sphere outside the allowed area - ``cell'' - is defined to be invalid. Modifying the radius of the spheres changes the allowed locations in the cell, which in turn affects the partition function of the system. The ``square'' refers to the piecewise intermolecular pair potential $\Phi_{12}$, shown mathematically in Eq. \ref{phi12} and graphically in Figure ~\ref{sw_phi}.

@@

The energy of the fluid is then always an integer multiple of the well depth $\epsilon$. This leads us to choose $\epsilon$ and $\sigma$ as natural scales for the quantities of our simulation. We then define dimensionless temperatures $T/\epsilon$, energies $E/\epsilon$, and distances $r/\sigma$. Moreover, we make the decision to work in geometric units for coding simplicity, setting the Boltzmann constant $k_B$ (as well as any other physical constants) to one. This then allows us to resonably compare energies to temperatures since they have the same units.

@@

Equations ~\ref{Fdef} and ~\ref{FdefSM} together bridge classical thermodynamics and statistical mechanics. Due to the nature of this project, we will primarily be concerned with the statistical mechanics definition (that is, Eq. ~\ref{FdefSM}) of the free energy. We are, then, ultimately concerned with determining the partition function for any system we are concerned with. The canonical ensemble partition function can be compactly represented by a summation:
\begin{align}
    Z &= \sum^{\text{energies}}~e^{-\beta E_i}
\end{align}
where $\beta =  T^{-1}$. Of course, it is typically impractical to sum over the individual energies of the system, especially if there are expected degeneracies. Instead we sum over states:
\begin{align}
    Z &= \sum^{\text{states}}~D(E)~e^{-\beta E_i}
\end{align}
where we have now introduced the density of states $D(E)$ which is a simple function that returns the number of states which have a given energy $E$. This introduction is now the focus of calculation - if we can accurately determine the density of states of the system and have knowledge of the available energies, we can calculate the Helmholtz free energy and, from there, any other  \\   
Since $F$ contains a natural log term, we can separate and characterize the individual contributions of kinetic energy (``ideal'') and energy associated with the configuration of the system (``excess'').  
A derivation of the two energy contributions for a monoatomic fluid follows. \\

@@

We begin with the canonical form of the canonical ensemble partition function $Z$:
\begin{align}
    Z &= \sum_i e^{-\beta E_i}
\end{align}
where we sum over individual energies of the system $E_i$. However, this method doesn't account for degeneracies in the system, so instead we sum over the states $s$ (to avoid including the density of states, which is difficult to find for most systems):
\begin{align}
    Z &= \sum_s e^{-\beta E_s}
\end{align}

@@

\begin{align}
    Z &= \int_s~ \diff\mathbf{r}~ \diff\mathbf{p} ~e^{-\beta V(\mathbf{r})}~e^{-\beta K(\mathbf{p})}
    %&= \sum_s e^{\beta V(\mathbf{r})}e^{-\beta T(\mathbf{p})}
\end{align} 
However, the above is only partially correct, as it is only for one particle. We can account for multiple particles by requiring that each particle has the same partition function contribution for kinetic energy; however, the potential energy does not seperate as nicely since it depends on the position of all particles. For the kinetic energy, we are simply integrating $N$ times for $N$ particles:

\begin{align}
    Z &= \int_{s_1} \int_{s_2} \dots \int_{s_N} \diff\mathbf{r_1}~ \diff\mathbf{p_1} ~\diff\mathbf{r_2} ~\diff\mathbf{p_2} ~\dots ~\diff\mathbf{r_N}~ \diff\mathbf{p_N} ~e^{-\beta V(\mathbf{r_1}, \mathbf{r_2}, \dots \mathbf{r_N})}~e^{-\beta(K(\mathbf{p_1}) + K(\mathbf{p_2}) ... K(\mathbf{p_N}))}\\
    &= \int_s \diff\mathbf{r}^N ~e^{-\beta V(\mathbf{r_1}, \mathbf{r_2}, \dots \mathbf{r_N})} \left[\int_s \diff\mathbf{p}~e^{-\beta K(\mathbf{p})}\right]^{N}
\end{align}
where we have separated the integrals by recognizing that the integrands are independant.\\
Additionally, for a proper quantum mechanical treatment we must assume that the particles are indistinguishable which means the above expression over-counts by a factor of $N!$. Our expression for the partition function is then

\begin{align}
    Z &= \frac{1}{N!}\int_s \diff\mathbf{r}^N ~e^{-\beta V(\mathbf{r_1}, \mathbf{r_2}, \dots \mathbf{r_N})} \left[\int_s \diff\mathbf{p}~e^{-\beta K(\mathbf{p})}\right]^{N}
    \label{Z_full}
\end{align}

@@

We begin by imagining a cell of length $L$ and volume $L^3$ in which lives a particle. Then, we quantize the allowed wavelengths of the particle within the cell by defining the wavevector $\mathbf{k}$ as
\begin{align}
    \mathbf{k} &= \frac{2\pi}{L}(n_x\mathbf{\hat{x}} + n_y\mathbf{\hat{y}} + n_z\mathbf{\hat{z}})
\end{align}
It should be noted that this definition of $k$ introduces three-fold degeneracy.\\
Remembering that kinetic energy can be expressed as $K = \frac{\hbar^2 k^2}{2m}$, and using our definition of $\mathbf{k}$:
\begin{align}
    Z_{id} &= \frac{1}{N!}\left[\int_s \diff\mathbf{p} ~e^{-\beta \frac{\hbar^2 k^2}{2m}}\right]^{N}
    \label{Zt_def}
\end{align}
This unfortunately introduces a mass quantity into the partition function which adds a parameter that needs fitting. [more] 
If we then allow the cell to become very large, then the we can easily see in Eq. \ref{Zt_def} that the spacing between occupants in phase space becomes small, so we can integrate instead $\diff ^3 n$ over all possible values of $n$:
\begin{align}
    Z_{id} &= \frac{1}{N!}\left[\int_0^\infty \diff^3 n ~e^{-\frac{\beta \hbar^2}{2m} \left(\frac{2\pi}{L}(n_x\mathbf{\hat{x}} + n_y\mathbf{\hat{y}} + n_z\mathbf{\hat{z}}\right)^2}\right]^{N} \\
    &= \frac{1}{N!}\left[\int_0^\infty \diff^3 n ~e^{ \frac{-2\beta \hbar^2\pi^2}{mL^2}(n_x^2 + n_y^2 + n_z^2)}\right]^{N}
\end{align}
Evaluating this integral yields
\begin{align}
    Z_{id} &= \left(\frac{V^N}{N!\Lambda^{3N}}\right) \,, \\
    \Lambda &= \frac{\hbar\sqrt{2\pi\beta}}{\sqrt{m}}
\end{align}
where we have introduced the thermal deBroglie wavelength $\Lambda$, loosely interpreted as [physical thing]. The full partition function for any monoatomic system is then
\begin{align}
    Z &= Z_{exc} \left(\frac{V}{\Lambda^{3}}\right)^{N} \,.
\end{align}
For liquids, the $Z_{exc}$ term tends to be difficult to evaluate; the hard sphere liquid, the simplest model for a fluid, does not have an analytic representation for the excess free energy. It follows that a considerable amount of work is required to determine the excess free energy for a given system.

@@

\section{Methods}
This work requires the intersection of three methods to calculate the absolute free energy $F^{abs}_{tot}$ of the bulk cell. A preliminary value of the free energy of the bulk can be obtained from Monte-Carlo simulations. This value can then be made absolute by carefully comparing the results with Thermodynamic Integration, a method of determining the excess free energy of a system. [more !fixme] 
\subsection{Monte-Carlo Methods}
Monte-Carlo methods utilize random or pseudorandom sampling and check for certain conditions or measure certain quantities after a specified number of iterations. Traditional random sampling requires that the current step, or ``move'', is irrespective of previous steps. True random movement is an ideal simulation characteristic for the canonical ``drunken rambler'' scenario, in which a intoxicated man attempts to make his way home by taking random steps. Since the drunken man chooses his next move arbitrarily, his movement is considered ``random''; it follows that true random sampling is desirable for this physical system. Statistical mechanics, however, does not exhibit this independence. Each time the man makes a move there is an equal probability that he will move in any particular direction. Rather than all movements having equal probability, statistical mechanics stipulates that the probability of a particular move occurring depends explicitly upon the energy difference of the current and resulting configurations. Instead of traditional Monte Carlo methods, Markov chain (specifically the Metropolis-Hastings algorithm) Monte Carlo is the natural decision.
\subsubsection{Markov Chain}
Markov chain methods eschew the true randomness of regular Monte Carlo by taking the current configuration into account before moving. If we define the initial state of the system by a vector column $\mathbf{q}$ (where the components refer to general coordinates, typically taken to be position of each sphere), and the resulting state $\mathbf{p}$ after a single time step, the relationship
\begin{align*}
    \mathbf{q} &= \mathcal{P}\, \mathbf{p}\\
\end{align*}
defines a ``transition'' of the system. $\mathcal{P}$ is a necessarily square matrix (as long as the number of coordinates remain fixed) that contains information about these transitions. We call this construction $\mathcal{P}$ the {\it transition matrix}. The transition matrix's $i^{\text{th}}$ eigenvalue the probability of the $i^{\text{th}}$ state occurring which, for the canonical ensemble. If all elements of the transition matrix are nonzero, we can see that any state of the system is reachable after a finite number of transitions from any given initial state. A physically judicious choice of transition matrix is one that satisfies the thermodynamic steady state condition; given the equilibrium state $\mathbf{q}_{eq}$, we enforce that
\begin{align*}
    \mathcal{P}\,\mathbf{q}_{eq} &= \mathbf{q}_{eq}
\end{align*}
The transition matrix thusly embodies the physics of the simulations [more?]

\ignore{\subsubsection{Histogram Methods}
Histograms are graphical representation of ``counting'' data. Monte Carlo histogram methods simply track the number of times the system has visited a particular energy. If the transition matrix 
\begin{figure}
    \caption{Sample histogram.}
\end{figure}}
\subsection{Broad Histogram Monte-Carlo}
If one records the energy of the system at each time step, it is straightforward to construct a histogram of energies.

@@

\subsection{Calculating Thermodynamic Quantities from Simulation Output}
Given the density of states output $\ln[D(E)]$ from the Monte-Carlo simulation, combined with {\it a priori} knowledge that the system obeys a canonical ensemble distribution of states, we can easily calculate, for instance, the excess internal energy:
\begin{align*}
     U_{exc}(T) &= \frac{\sum_{i=0}^{\text{\tiny Energies}}\,E_i\, \exp\left(\ln[D(E)] - \beta E_i\right)}{\sum_{i=0}^{\text{\tiny Energies}} \exp[\ln(D(E))] - \beta E_i}\\
     &= \frac{\sum_{i=0}^{\text{\tiny Energies}} E_i \exp\left(-\beta E_i\right)}{Z}
\end{align*} 
where, as usual, $\beta = T^{-1}$. Taking the exponential of $\ln(D(E))$, unfortunately, tends to introduce overflow or underflow errors during computation. We combat this by subtracting the maximum value of the natural log of the density of states \ignore{subtracting maximum ought to make it more negative\dots} from each element of $D(E)$, and then adding the maximum back in after the exponential. We calculate the excess free energy by using the entropy equality developed earlier:
\begin{align*}
     S_{SW,~exc}\Bigm\lvert_{T=\infty} &= S_{HS,~exc}\\
     \implies F_{SW,~exc} &= -TS_{HS,~exc} + F_{inf}\\
     F_{SW,~exc} &= -TS_{HS,~exc} - T\ln\left(\frac{Z}{Z_{inf}}\right)
\end{align*} 
The excess hard sphere entropy $T_{HS,~exc}$ is calculated directly from thermodynamic integration and both the partition functions are calculated the usual way [MORE ABOUT $Z_inf$]. Once we have calculated both the excess free energy and excess internal energy, we can directly find the excess entropy:
\begin{align*}
    S_{SW, ~exc} &= \frac{1}{T}\left(U_{SW,~exc} - F_{SW,~exc} \right)
\end{align*}

@@

\begin{figure}
\centering
    \includegraphics[width=.75\textwidth]{../renormalization/figs/Shs-vs-eta.pdf}
    \caption{Excess hard sphere entropy as a function of density.}
    \label{Shs-eta}
\end{figure}
\ignore{Filling fraction or density? Value in distinction? We never use number density for the sake of units...}
Figure \ref{Shs-eta} shows the computed excess entropy as a function of density. We see that the both the first and second doubling regime closely match the Carnahan-Starling prediction for the entropy. After freezing, the first doubling regime veers from the prediction by the Carnahan-Starling equation of state.\\

@@

\section{Conclusion}
We have presented a method by which the properties of bulk fluids can be estimated by decomposing the bulk into individual cells in a renormalization-minded manner by exploiting the critical behavior of fluids. The free energy contribution of each length scale can then be investigated and compared, both to determine the most energetic density fluctuations and to provide calculations that serve as a intermediate comparison with SAFT methods. This method is immediately applicable to any system where the energy of a configuration is straightforward to compute (that is, relatively inexpensive computationally). We presented an application for a system that obeys the canonical ensemble, but the extension to the grand canonical ensemble can be done with little modification.  
%\bibliographystyle{plain}
\bibliography{Thesis}
\end{document}

>>>>

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry}
\setlength{\parindent}{0cm} % don't indent new paragraphs...
%\parskip \psl % ... place a space between paragraphs instead
\usepackage{graphicx, float}
\usepackage[font=small, labelfont=bf]{caption}
%\usepackage[english]{babel}
\usepackage{tikz}
%\usepackage{biblatex}
\usepackage{amsmath}
\usepackage[inline]{enumitem}
%\usepackage{float}

% Header definition
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead{}
%\renewcommand{\sectionmark}[1]{\markright{\thesection.\ #1}}

\fancyhead[LE, RO]{\slshape \rightmark}
\fancyhead[RE, LO]{\slshape Chapter \leftmark}

% end header

\usetikzlibrary{automata, positioning}
\newcommand{\ignore}[1]{}
\newcommand*{\diff}{\mathrm{d}}
\newcommand{\some}{{\color{red} something}}
\renewcommand{\vec}[1]{\mathbf{#1}}

@@

\renewcommand{\title}{Free Energy Decomposition of the Critical Square-Well Liquid Using a Renormalization Group Method}

@@

\section{Introduction}
\subsection{Motivation}
Fluids have long garnered interest\cite{theorysimpleliquids}\cite{theoryofcrits}, particularly among computationally-minded researchers, as an area in which current models are unequipped to robustly predict behavior across multiple regimes. A largely troublesome \cite{white2001global} regime is near the so-called ``critical point'' where liquid and vapor become indistinguishable and, more importantly, large fluctuations of fluid density occur. Most traditional fluid models assume the density across the cell to be nearly constant. As a result, these methods tend to poorly handle these fluctuations near criticality. 
Using the well-established model of the square-well fluid, we seek to determine the applicability of a particular generalized renormalization-group formalism (``GRG'') to find the absolute Helmholtz free energy of systems near criticality.\\

Past efforts to incorporate renormalization group methods into standard Statistical Associating Fluid Theory (``SAFT'') methods have seen success [cite me], particularly the treatment offered by Forte et al.~\cite{forte2011application} with the inclusion of a fitted parameter related to the correlation length. GRG's history in fluid simulation has largely been a fitted one - that is, determining a free parameter and then simulating a vast number of cells, varying the parameter each time, to find the most well-suited value for a particular fluid [!fixme not quite]. We present a non-fitted approach that is GRG-minded from the ground up. We exploit the periodicity of Monte-Carlo methods to decompose large cells in the thermodynamic limit into small cells that only allow certain fluctuation wavelengths. We then carefully couple simulations of different ``doubling regimes'', each of which considers a different maximum wavelength of fluctuations, where the number density remains fixed but the volume and number have increased. Summing the difference in absolute Helmholtz free energy $\Delta F^{\text{abs}}$ between doubling regimes will then approximate the absolute free energy of the full cell, including near criticality. Finding absolute free energies is a somewhat challenging problem, but is certainly surmountable with more Monte Carlo. \\
We will begin by outlining the necessary models and assumptions/approximations for this work. We will then detail the algorithms used to compute the quantities with the the discussion on results following. 
\subsection{Hard Sphere}
The hard sphere model approximates monoatomic fluids as having no interactions other than collisions. These spheres exhibit no attraction, and consequently each carries kinetic energy $K = 3/2 k_b T$. There is also a harder-to-calculate [!fixme] ``excess'' energy to be considered (which, in fact, contains most of the interesting behavior) which will be discussed further on. The system is defined to have zero total linear momentum, and any collision between two or more spheres is perfectly elastic. For simulation purposes, the system is required to have periodic boundary conditions - that is, when a sphere would classically collide with the wall of the cell it instead breaches the wall and then reappears on the other side of the cell with the same velocity. \\
\ignore{The hard sphere model is incredibly useful; as high temperatures are approached, energy associated with sphere location tends to asymptote for most models [are there any that this isn't true for?] and the kinetic term consequently dominates.}

\subsection{Square-Well Liquid}
The square-well liquid is a common approximation that sees extensive use throughout computational fluid research. This simple fluid model consists of hard spheres that are subject to an intermolecular pair potential - that is, an attractive force between spheres. This model consists of several key parameters: sphere size $\sigma$, well width $\lambda$, and well depth $\epsilon$. \\
As a hard-sphere model, the spheres are not allowed to overlap. Any arrangement of spheres - hereafter referred to as the ``state'' of the system - that involves either two or more spheres overlapping or a sphere outside the allowed area - ``cell'' - is defined to be invalid. Modifying the radius of the spheres changes the allowed locations in the cell, which in turn affects the partition function of the system. The ``square'' refers to the piecewise intermolecular pair potential $\Phi_{12}$, shown mathematically in Eq. \ref{phi12} and graphically in Figure ~\ref{sw_phi}.

@@

The potential energy of the fluid is then always an integer multiple of the well depth $\epsilon$. This leads us to choose $\epsilon$ and $\sigma$ as natural scales for the quantities of our simulation. We then define dimensionless temperatures $T/\epsilon$, energies $E/\epsilon$, and distances $r/\sigma$. Moreover, we make the decision to work in geometric units for coding simplicity, setting the Boltzmann constant $k_B$ (as well as any other physical constants) to one. This then allows us to resonably compare energies to temperatures since they have the same units.

@@

Equations ~\ref{Fdef} and ~\ref{FdefSM} together bridge classical thermodynamics and statistical mechanics. Due to the nature of this project, we will primarily be concerned with the statistical mechanics definition (that is, Eq. ~\ref{FdefSM}) of the free energy. We are, then, ultimately concerned with determining the partition function for any system we are concerned with. The canonical ensemble partition function can be compactly represented by a summation:
\begin{align}
    Z &= \sum^{\text{states}}~e^{-\beta E_i}
\end{align}
where $\beta =  T^{-1}$ and $E_i$ is the energy of the $i^{\text{th}}$. Of course, it is typically impractical to sum over the individual states of the system, especially if there are expected degeneracies. Instead we sum over energies:
\begin{align}
    Z &= \sum^{\text{energies}}~D(E)~e^{-\beta E_i}
\end{align}
where we have now introduced the density of states $D(E)$ which is a simple function that returns the number of states which have a given energy $E$. This introduction is now the focus of calculation - if we can accurately determine the density of states of the system and have knowledge of the available energies, we can calculate the Helmholtz free energy and, from there, any other  \\   
Since $F$ contains a natural log term, we can separate and characterize the individual contributions of kinetic energy (``ideal'') and energy associated with the configuration of the system (``excess'').  
A derivation of the two energy contributions for a monoatomic fluid follows. \\

@@

We begin with the canonical form of the canonical ensemble partition function $Z$:
\begin{align}
    Z &= \sum_i D(E) e^{-\beta E_i}
\end{align}
where we sum over individual energies of the system $E_i$. [reason that we sum over states] , so instead we sum over the states $s$:
\begin{align}
    Z &= \sum_s e^{-\beta E_s}
\end{align}

@@

\begin{align}
    Z &= \int_s~ \diff\mathbf{r}~ \diff\mathbf{p} ~e^{-\beta V(\mathbf{r})}~e^{-\beta K(\mathbf{p})}
    %&= \sum_s e^{\beta V(\mathbf{r})}e^{-\beta T(\mathbf{p})}
\end{align} 
However, the above is only partially correct, as it is only for one particle. We can account for multiple particles by requiring that each particle has the same partition function contribution for kinetic energy; however, the potential energy does not separate as nicely since it depends on the position of all particles. For the kinetic energy, we are simply integrating $N$ times for $N$ particles:

\begin{align}
    Z &= \int_{s_1} \int_{s_2} \dots \int_{s_N} \diff\mathbf{r_1}~ \diff\mathbf{p_1} ~\diff\mathbf{r_2} ~\diff\mathbf{p_2} ~\dots ~\diff\mathbf{r_N}~ \diff\mathbf{p_N} ~e^{-\beta V(\mathbf{r_1}, \mathbf{r_2}, \dots \mathbf{r_N})}~e^{-\beta(K(\mathbf{p_1}) + K(\mathbf{p_2}) ... K(\mathbf{p_N}))}\\
    &= \int_s \diff\mathbf{r}^N ~e^{-\beta V(\mathbf{r_1}, \mathbf{r_2}, \dots \mathbf{r_N})} \left[\int_s \diff\mathbf{p}~e^{-\beta K(\mathbf{p})}\right]^{N}
    \label{Z-intermed}
\end{align}
where we have separated the integrals by recognizing that the integrands are independent.\\
Additionally, for a proper quantum mechanical treatment we must assume that the particles are indistinguishable which means the above expression over-counts by a factor of $N!$. Additionally, we will multiply Eq. \ref{Z-intermed} by an overall factor of $\frac{V^N}{V^N}$ to preserve units and extensivity. Our expression for the partition function is then

\begin{align}
    Z &= \frac{1}{V^N N!}\int_s \diff\mathbf{r}^N ~e^{-\beta V(\mathbf{r_1}, \mathbf{r_2}, \dots \mathbf{r_N})} \left[V \int_s \diff\mathbf{p}~e^{-\beta K(\mathbf{p})}\right]^{N}
    \label{Z_full}
\end{align}

@@

We begin by imagining a cell of length $L$ and volume $L^3$ in which lives a particle. Then, we quantize the allowed wavelengths of the particle within the cell by defining the wavevector $\mathbf{k}$ as
\begin{align}
    \mathbf{k} &= \frac{2\pi}{L}(n_x\mathbf{\hat{x}} + n_y\mathbf{\hat{y}} + n_z\mathbf{\hat{z}})
\end{align}
It should be noted that this definition of $k$ introduces three-fold degeneracy.\\
Remembering that kinetic energy can be expressed as $K = \frac{\hbar^2 k^2}{2m}$, and using our definition of $\mathbf{k}$:
\begin{align}
    Z_{id} &= \frac{1}{N!}\left[\int_s \diff\mathbf{p} ~e^{-\beta \frac{\hbar^2 k^2}{2m}}\right]^{N}
    \label{Zt_def}
\end{align}
This unfortunately introduces a mass quantity into the partition function which adds an additional parameter that needs to be determined. [more] 
If we then allow the cell to become very large, then the we can easily see in Eq. \ref{Zt_def} that the spacing between occupants in phase space becomes small, so we can integrate instead $\diff ^3 n$ over all possible values of $n$:
\begin{align}
    Z_{id} &= \frac{1}{N!}\left[\int_0^\infty \diff^3 n ~e^{-\frac{\beta \hbar^2}{2m} \left(\frac{2\pi}{L}(n_x\mathbf{\hat{x}} + n_y\mathbf{\hat{y}} + n_z\mathbf{\hat{z}}\right)^2}\right]^{N} \\
    &= \frac{1}{N!}\left[\int_0^\infty \diff^3 n ~e^{ \frac{-2\beta \hbar^2\pi^2}{mL^2}(n_x^2 + n_y^2 + n_z^2)}\right]^{N}
\end{align}
Evaluating this integral yields
\begin{align}
    Z_{id} &= \left(\frac{V^N}{N!\Lambda^{3N}}\right) \,, \\
    \Lambda &= \frac{\hbar\sqrt{2\pi\beta}}{\sqrt{m}}
\end{align}
where we have introduced the thermal deBroglie wavelength $\Lambda$ which defines the de Broglie wavelength of an ideal gas at a temperature $T$. The full partition function for any monoatomic system is then
\begin{align}
    Z &= Z_{exc} \left(\frac{V}{\Lambda^{3}}\right)^{N} \,.
\end{align}
For liquids [fluids?], the $Z_{exc}$ term is difficult to ever evaluate; the hard sphere liquid, the simplest model for a dense fluid, does not have an analytic representation for the excess free energy. It follows that a considerable amount of work is required to determine the excess free energy for a given system.

@@

\section{Methods}
This work requires the intersection of three methods to calculate the absolute free energy $F^{abs}_{tot}$ of the bulk cell. A preliminary value of the free energy of the bulk can be obtained from Monte-Carlo simulations. This value can then be made absolute by carefully comparing the results with Thermodynamic Integration, a method of determining the excess free energy of a system. [more !fixme] 
\subsection{Monte-Carlo Methods}
Monte-Carlo methods utilize random or pseudorandom sampling and check for certain conditions or measure certain quantities after a specified number of iterations. Traditional random sampling requires that the current step, or ``move'', is irrespective of previous steps. True random movement is an ideal simulation characteristic for the canonical ``drunken rambler'' scenario, in which a intoxicated man attempts to make his way home by taking random steps. Since the drunken man chooses his next move arbitrarily, his movement is considered ``random''; it follows that true random sampling is desirable for this physical system. Statistical mechanics, however, does not exhibit this independence. Each time the drunken man makes a move there is an equal probability that he will move in any particular direction. Rather than all movements having equal probability, statistical mechanics stipulates that the probability of a particular move occurring depends explicitly upon the energy difference of the current and resulting configurations. Instead of true random Monte Carlo methods, we clearly need some way to control the probability of transitions from a given state. In pursuit of this control, we employ the method of Markov chains, specifically Metropolis-Hastings Monte Carlo.
\subsubsection{Markov Chain}
[cite me pls]
Markov chain methods eschew the true randomness of regular Monte Carlo by taking the current configuration into account before moving. If we define the initial state of the system by a vector column $\mathbf{q}$ (where the components refer to general coordinates, typically taken to be position of each sphere), and the resulting state $\mathbf{p}$ after a single time step, the relationship
\begin{align*}
    \mathbf{p} &= \mathcal{P}\, \mathbf{q}\\
\end{align*}
defines a ``transition'' of the system. $\mathcal{P}$ is a necessarily square matrix (as long as the number of coordinates remain fixed) that contains information about these transitions. We call this construction $\mathcal{P}$ the {\it transition matrix}. The transition matrix's $i^{\text{th}}$ eigenvalue the probability of the $i^{\text{th}}$ state occurring which, for the canonical ensemble. If all elements of the transition matrix are nonzero, we can see that any state of the system is reachable after a finite number of transitions from any given initial state. [This is the best I could synthesize Hansen, following statement could be wrong!] A physically judicious choice of transition matrix is one that satisfies the thermodynamic steady state condition; given the equilibrium state $\mathbf{q}_{eq}$, we enforce that
\begin{align*}
    \mathcal{P}\,\mathbf{q}_{eq} &= \mathbf{q}_{eq}
\end{align*}
The transition matrix thus embodies the physics of the simulation. [more?]

\ignore{\subsubsection{Histogram Methods}
Histograms are graphical representation of ``counting'' data. Monte Carlo histogram methods simply track the number of times the system has visited a particular energy. If the transition matrix 
\begin{figure}
    \caption{Sample histogram.}
\end{figure}}
\subsection{Broad Histogram Monte-Carlo}
If one records the energy of the system at each time step, it is straightforward to construct a histogram of energies. Typical implementations of histogram methods exhibit histograms that sharply peak near a central value $E_0$ which then symmetrically decays. Of course, at higher temperatures, the histogram broadens and flattens out as more energies become easily accessible and, in the limit that temperature becomes infinite, the histogram becomes completely flat since all states are equally likely. If the simulation is allowed to run for an infinite amount of time, the histogram will be precisely detail the system. x The tails of the histogram, however, contain spurious statistics because the system rarely visits these energies. Gathering sufficient statistics

@@

\subsection{Calculating Thermodynamic Quantities from Simulation Output}
Given the density of states output $\ln[D(E)]$ from the Monte-Carlo simulation, combined with {\it a priori} knowledge that the system obeys a canonical ensemble distribution of states, we can easily calculate, for instance, the excess internal energy:
\begin{align*}
     U_{exc}(T) &= \frac{\sum_{i=0}^{\text{\tiny Energies}}\,E_i\, \exp\left(\ln[D(E)] - \beta E_i\right)}{\sum_{i=0}^{\text{\tiny Energies}} \exp[\ln(D(E))] - \beta E_i}\\
     &= \frac{\sum_{i=0}^{\text{\tiny Energies}} E_i \exp\left(-\beta E_i\right)}{Z}
\end{align*} 
where, as usual, $\beta = T^{-1}$. Taking the exponential of $\ln(D(E))$, unfortunately, tends to introduce overflow or underflow errors during computation. We combat this by subtracting the maximum value of the natural log of the density of states \ignore{subtracting maximum ought to make it more negative\dots} from each element of $D(E)$, and then adding the maximum back in after the exponential. We calculate the excess free energy by using the entropy equality developed previously:
\begin{align*}
     S_{SW,~exc}\Bigm\lvert_{T=\infty} &= S_{HS,~exc}\\
     \implies F_{SW,~exc} &= -TS_{HS,~exc} + F_{inf}\\
     F_{SW,~exc} &= -TS_{HS,~exc} - T\ln\left(\frac{Z}{Z_{inf}}\right)
\end{align*} 
The excess hard sphere entropy $T_{HS,~exc}$ is calculated directly from thermodynamic integration and both the partition functions are calculated the usual way [MORE ABOUT $Z_inf$]. Once we have calculated both the excess free energy and excess internal energy, we can directly find the excess entropy:
\begin{align*}
    S_{SW, ~exc} &= \frac{1}{T}\left(U_{SW,~exc} - F_{SW,~exc} \right)
\end{align*}

@@

\begin{figure}
\centering
    \includegraphics[width=.75\textwidth]{../renormalization/figs/Shs-vs-eta.pdf}
    \caption{Excess hard sphere entropy as a function of density.}
    \label{Shs-eta}
\end{figure}
\ignore{Filling fraction or density? Value in distinction? We never use number density for the sake of units...}
Figure \ref{Shs-eta} shows the computed excess entropy as a function of density. We see that the both the first and second doubling regime closely match the Carnahan-Starling prediction for the entropy. After freezing, the first doubling regime veers from the prediction by the Carnahan-Starling equation of state; since the Carhan-Starling approximation is accurate only for low density fluids, we expect this behavior.\\

@@

\section{Conclusion}
We have presented a method by which the properties of bulk fluids can be estimated by decomposing the bulk into individual cells in a renormalization-minded manner by exploiting the critical behavior of fluids. The free energy contribution of each length scale can then be investigated and compared, both to determine the most energetic density fluctuations and to provide calculations that serve as a intermediate comparison with SAFT methods. This method is immediately applicable to any system where the energy of a configuration is straightforward to compute (that is, relatively inexpensive computationally). The application presented was a system that obeys the canonical ensemble, but the extension to the grand canonical ensemble can be done with little modification. Future work can include stuff [more]. 
%\bibliographystyle{plain}
\cleardoublepage
\bibliography{Thesis}
\end{document}
